---
title: "India Census 2011: Final Presentation Report"
author: "Vivek Reddy"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 4
    number_sections: true
    theme: flatly
    css: final-report.css
    fig_width: 11
    fig_height: 6
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 11,
  fig.height = 6,
  dpi = 150
)
```

```{r libs, include=FALSE}
required_packages <- c(
  "readxl", "dplyr", "tidyr", "ggplot2", "scales", "forcats",
  "viridis", "kableExtra", "corrplot", "tibble"
)

new_packages <- required_packages[!(required_packages %in% rownames(installed.packages()))]
if (length(new_packages) > 0) {
  install.packages(new_packages, repos = "https://cran.r-project.org")
}

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(forcats)
library(viridis)
library(kableExtra)
library(corrplot)
library(tibble)
```

```{r theme, include=FALSE}
theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "#4a4a4a"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      axis.text.x = element_text(angle = 35, hjust = 1)
    )
)
```

<div class="hero">
<h1>India Census 2011</h1>
<p>State and District Demographic Intelligence Dashboard</p>
<p class="hero-sub">A presentation-ready statistical brief focused on population, gender balance, literacy, social composition, and labor structure.</p>
</div>

```{r data, include=FALSE}
raw_data <- read_excel("2011-IndiaStateDist.xlsx", sheet = 1)

state_total <- raw_data %>%
  filter(Level == "STATE", TRU == "Total") %>%
  mutate(
    Sex_Ratio = round((TOT_F / TOT_M) * 1000, 1),
    Child_Sex_Ratio = round((F_06 / M_06) * 1000, 1),
    Literacy_Rate = round((P_LIT / (TOT_P - P_06)) * 100, 2),
    Male_Literacy_Rate = round((M_LIT / (TOT_M - M_06)) * 100, 2),
    Female_Literacy_Rate = round((F_LIT / (TOT_F - F_06)) * 100, 2),
    Literacy_Gender_Gap = round(Male_Literacy_Rate - Female_Literacy_Rate, 2),
    Pct_SC = round((P_SC / TOT_P) * 100, 2),
    Pct_ST = round((P_ST / TOT_P) * 100, 2),
    Pct_Child_06 = round((P_06 / TOT_P) * 100, 2),
    Work_Participation_Rate = round((TOT_WORK_P / TOT_P) * 100, 2),
    Female_WPR = round((TOT_WORK_F / TOT_F) * 100, 2),
    Pct_Cultivators = round(((MAIN_CL_P + MARG_CL_P) / TOT_WORK_P) * 100, 2),
    Pct_Agri_Lab = round(((MAIN_AL_P + MARG_AL_P) / TOT_WORK_P) * 100, 2),
    Pct_Other_Workers = round(((MAIN_OT_P + MARG_OT_P) / TOT_WORK_P) * 100, 2)
  )

state_urban <- raw_data %>%
  filter(Level == "STATE", TRU == "Urban") %>%
  select(Name, Urban_Pop = TOT_P)

state_total <- state_total %>%
  left_join(state_urban, by = "Name") %>%
  mutate(Urban_Pct = round((Urban_Pop / TOT_P) * 100, 2))

district_total <- raw_data %>%
  filter(Level == "DISTRICT", TRU == "Total") %>%
  mutate(
    Sex_Ratio = round((TOT_F / TOT_M) * 1000, 1),
    Child_Sex_Ratio = round((F_06 / M_06) * 1000, 1),
    Literacy_Rate = round((P_LIT / (TOT_P - P_06)) * 100, 2),
    Work_Participation_Rate = round((TOT_WORK_P / TOT_P) * 100, 2)
  )
```

```{r kpis, results='asis'}
india_pop <- sum(state_total$TOT_P, na.rm = TRUE)
india_lit <- sum(state_total$P_LIT, na.rm = TRUE) / sum(state_total$TOT_P - state_total$P_06, na.rm = TRUE) * 100
india_sex_ratio <- sum(state_total$TOT_F, na.rm = TRUE) / sum(state_total$TOT_M, na.rm = TRUE) * 1000
india_wpr <- sum(state_total$TOT_WORK_P, na.rm = TRUE) / sum(state_total$TOT_P, na.rm = TRUE) * 100

cat("<div class='kpi-grid'>")
cat(sprintf("<div class='kpi'><div class='kpi-title'>Total Population</div><div class='kpi-value'>%s</div></div>", comma(india_pop)))
cat(sprintf("<div class='kpi'><div class='kpi-title'>Literacy Rate</div><div class='kpi-value'>%.1f%%</div></div>", india_lit))
cat(sprintf("<div class='kpi'><div class='kpi-title'>Sex Ratio</div><div class='kpi-value'>%.0f</div></div>", india_sex_ratio))
cat(sprintf("<div class='kpi'><div class='kpi-title'>Work Participation</div><div class='kpi-value'>%.1f%%</div></div>", india_wpr))
cat("</div>")
```

## 1. Population Footprint

```{r pop-state}
top12 <- state_total %>%
  arrange(desc(TOT_P)) %>%
  head(12) %>%
  mutate(Name = fct_reorder(Name, TOT_P))

ggplot(top12, aes(Name, TOT_P / 1e7, fill = TOT_P)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_viridis_c(option = "C") +
  labs(
    title = "Top 12 States/UTs by Population",
    subtitle = "Population shown in Crores",
    x = NULL,
    y = "Population (Crores)"
  )
```

```{r pop-bubble}
ggplot(state_total, aes(x = Literacy_Rate, y = Urban_Pct, size = TOT_P, color = Sex_Ratio)) +
  geom_point(alpha = 0.75) +
  geom_text(aes(label = Name), check_overlap = TRUE, size = 3, vjust = -0.8) +
  scale_size_continuous(range = c(2, 20), guide = "none") +
  scale_color_viridis_c(option = "C") +
  labs(
    title = "State Demography Bubble Map",
    subtitle = "X = Literacy, Y = Urbanization, Bubble size = Population, Color = Sex Ratio",
    x = "Literacy Rate (%)",
    y = "Urban Population (%)",
    color = "Sex Ratio"
  )
```

## 2. Gender Balance and Child Profile

```{r gender-bars}
gender_df <- state_total %>%
  select(Name, Sex_Ratio, Child_Sex_Ratio) %>%
  pivot_longer(cols = c(Sex_Ratio, Child_Sex_Ratio), names_to = "Metric", values_to = "Value") %>%
  mutate(
    Metric = ifelse(Metric == "Sex_Ratio", "Overall Sex Ratio", "Child Sex Ratio (0-6)")
  )

ggplot(gender_df, aes(x = fct_reorder(Name, Value, .fun = mean), y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  coord_flip() +
  geom_hline(yintercept = 1000, linetype = "dashed", color = "#333333") +
  scale_fill_manual(values = c("Overall Sex Ratio" = "#0f766e", "Child Sex Ratio (0-6)" = "#be123c")) +
  labs(
    title = "Overall vs Child Sex Ratio by State",
    subtitle = "Dashed line indicates parity at 1000",
    x = NULL,
    y = "Females per 1000 Males",
    fill = NULL
  )
```

```{r child-share}
ggplot(state_total, aes(x = fct_reorder(Name, Pct_Child_06), y = Pct_Child_06, fill = Pct_Child_06)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "#fde68a", high = "#b45309") +
  labs(
    title = "Share of Children (0-6 years) in Population",
    x = NULL,
    y = "Child Population (%)"
  )
```

## 3. Literacy and Human Capital

```{r literacy-state}
ggplot(state_total, aes(x = fct_reorder(Name, Literacy_Rate), y = Literacy_Rate, fill = Literacy_Rate)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "#bfdbfe", high = "#1d4ed8") +
  labs(
    title = "Literacy Rate by State",
    x = NULL,
    y = "Literacy Rate (%)"
  )
```

```{r literacy-gap}
ggplot(state_total, aes(x = Female_Literacy_Rate, y = Male_Literacy_Rate, size = TOT_P, color = Urban_Pct)) +
  geom_point(alpha = 0.8) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#444444") +
  scale_size_continuous(range = c(2, 12), guide = "none") +
  scale_color_viridis_c(option = "B") +
  labs(
    title = "Female vs Male Literacy",
    subtitle = "Diagonal parity line; color indicates urbanization",
    x = "Female Literacy Rate (%)",
    y = "Male Literacy Rate (%)",
    color = "Urban %"
  )
```

## 4. Social Composition (SC/ST)

```{r scst}
scst <- state_total %>%
  select(Name, Pct_SC, Pct_ST) %>%
  pivot_longer(cols = c(Pct_SC, Pct_ST), names_to = "Group", values_to = "Pct") %>%
  mutate(Group = ifelse(Group == "Pct_SC", "Scheduled Caste", "Scheduled Tribe"))

ggplot(scst, aes(x = fct_reorder(Name, Pct, .fun = sum), y = Pct, fill = Group)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Scheduled Caste" = "#1d4ed8", "Scheduled Tribe" = "#f97316")) +
  labs(
    title = "SC/ST Composition by State",
    x = NULL,
    y = "Population Share (%)",
    fill = NULL
  )
```

## 5. Work and Economic Structure

```{r workforce}
workforce <- state_total %>%
  select(Name, Work_Participation_Rate, Female_WPR, Pct_Cultivators, Pct_Agri_Lab, Pct_Other_Workers)

p1 <- ggplot(workforce, aes(fct_reorder(Name, Work_Participation_Rate), Work_Participation_Rate, fill = Work_Participation_Rate)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "#bbf7d0", high = "#166534") +
  labs(title = "Work Participation Rate", x = NULL, y = "%")

p2 <- ggplot(workforce, aes(fct_reorder(Name, Female_WPR), Female_WPR, fill = Female_WPR)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "#fbcfe8", high = "#9d174d") +
  labs(title = "Female Work Participation", x = NULL, y = "%")

print(p1)
print(p2)
```

```{r sector-mix}
sectors <- state_total %>%
  select(Name, Pct_Cultivators, Pct_Agri_Lab, Pct_Other_Workers) %>%
  pivot_longer(cols = -Name, names_to = "Sector", values_to = "Pct") %>%
  mutate(
    Sector = recode(
      Sector,
      "Pct_Cultivators" = "Cultivators",
      "Pct_Agri_Lab" = "Agricultural Labour",
      "Pct_Other_Workers" = "Other Workers"
    )
  )

ggplot(sectors, aes(x = fct_reorder(Name, Pct, .fun = sum), y = Pct, fill = Sector)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Cultivators" = "#65a30d", "Agricultural Labour" = "#f59e0b", "Other Workers" = "#0ea5e9")) +
  labs(
    title = "Employment Sector Mix by State",
    x = NULL,
    y = "Worker Share (%)",
    fill = NULL
  )
```

## 6. District-Level Distribution Lens

```{r district-distributions}
pop_dist <- ggplot(district_total, aes(TOT_P / 1e6)) +
  geom_histogram(bins = 45, fill = "#0ea5e9", color = "white") +
  labs(title = "District Population Distribution", x = "Population (Millions)", y = "District Count")

lit_dist <- ggplot(district_total, aes(Literacy_Rate)) +
  geom_histogram(bins = 40, fill = "#6366f1", color = "white") +
  labs(title = "District Literacy Distribution", x = "Literacy Rate (%)", y = "District Count")

print(pop_dist)
print(lit_dist)
```

## 7. Correlation Structure

```{r corr, fig.width=12, fig.height=10}
corr_data <- state_total %>%
  select(
    Sex_Ratio, Child_Sex_Ratio, Literacy_Rate, Female_Literacy_Rate,
    Pct_SC, Pct_ST, Pct_Child_06, Work_Participation_Rate,
    Female_WPR, Urban_Pct
  )

# Drop columns that are entirely missing or have zero variance to avoid hclust/corrplot failures
usable_cols <- sapply(corr_data, function(x) sum(!is.na(x)) > 2 && stats::sd(x, na.rm = TRUE) > 0)
corr_data <- corr_data[, usable_cols, drop = FALSE]

if (ncol(corr_data) >= 2) {
  corr_mat <- cor(corr_data, use = "pairwise.complete.obs")
  corrplot(
    corr_mat,
    method = "color",
    type = "lower",
    order = "hclust",
    tl.col = "black",
    addCoef.col = "black",
    number.cex = 0.6,
    tl.srt = 30,
    col = colorRampPalette(c("#b91c1c", "white", "#1d4ed8"))(200)
  )
} else {
  plot.new()
  text(0.5, 0.5, "Insufficient variation for correlation matrix", cex = 1.2)
}
```

## 8. Final Snapshot Table

```{r table}
state_total %>%
  transmute(
    `State/UT` = Name,
    `Population` = comma(TOT_P),
    `Sex Ratio` = Sex_Ratio,
    `Literacy %` = round(Literacy_Rate, 1),
    `Female Literacy %` = round(Female_Literacy_Rate, 1),
    `WPR %` = round(Work_Participation_Rate, 1),
    `Urban %` = round(Urban_Pct, 1)
  ) %>%
  arrange(desc(as.numeric(gsub(",", "", Population)))) %>%
  kbl(caption = "State/UT Comparative Snapshot", align = "lrrrrrr") %>%
  kable_classic(full_width = FALSE)
```

<div class="footer-note">
This final page is deployment-oriented: visual-first, narrative-driven, and code-hidden for stakeholder presentation.
</div>
